---
title: "From raw logger to reconstructed track: the full GLS pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From raw logger to reconstructed track: the full GLS pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Why a second stage?

The calibration workflow in `glscalibrator` (v0.1) returns *threshold* positions:
longitude is well constrained, but latitude is derived from day length and is
noise-dominated — enough to place a coastal bird thousands of km off its true
latitude. The v0.2 pipeline fixes this by constraining latitude with the
logger's **own temperature**, matched to satellite SST, together with a movement
prior and a land mask, using the `probGLS` particle algorithm
(Merkel et al. 2016).

The pipeline is deliberately **taxon-agnostic**: everything biological lives in a
configuration object, not in the code.

## 1. Configure (once per study)

```{r}
library(glscalibrator)

# Start from the shipped template, or build in code:
cfg <- probgls_config(
  tagging_location = c(-115.174, 27.852),   # c(lon, lat) of the colony  <-- EDIT
  boundary_box     = c(-140, -80, 0, 55),   # plausible movement bounds   <-- EDIT
  NOAA_OI_location = "results/oisst",       # environmental-data folder
  speed_wet        = c(1, 1.3, 5),          # general seabird defaults
  speed_dry        = c(12, 6, 45)
)
cfg                       # printed nicely; run-time fields show as <derived>
write_probgls_config(cfg, "config.yml")     # share / version-control it
```

### Starting from a collaborator's parameter sheet

Movement-model settings often arrive as a spreadsheet — one row per
`prob_algorithm` argument, one column per study. Read it straight in:

```{r}
cfg <- read_probgls_params("variables_prob_gls.xlsx")   # or a .csv
```

Blank rows and formula rows (`min(trn$tFirst)`) keep the package defaults, and
two common transcription slips are repaired with a warning: a `tagging.location`
written (lat, lon), and a `boundary.box` written in interleaved order. The
deployment site is used as the decisive test — a box that does not contain its
own colony is not the intended box. Always read the warnings and confirm the
result:

```{r}
cfg[["tagging.location"]]
cfg[["boundary.box"]]
```

## 2. Get the environmental fields (once per date span)

```{r}
get_environmental_data(
  dates = as.Date(c("2023-06-01", "2024-05-15")),
  dest  = "results/oisst"
)
```

## 3. Run one deployment

```{r}
out <- run_gls_pipeline("birds/BW148", cfg)
head(out$result$track)                      # datetime, lon, lat (+ CIs)
plot_probgls_track(out, colony = cfg$tagging.location)
```

Under the hood `run_gls_pipeline()` calls, in order:
`read_gls()` -> `detect_twilights()` -> `filter_twilights()` -> `prepare_trn()`
-> `deduce_sst()` -> `run_probgls()`.

## 4. Run a whole colony

```{r}
batch <- run_gls_pipeline_batch(
  root       = "birds",
  config     = cfg,
  output_dir = "results/probGLS"
)
batch$summary
summarise_tracks(batch, lat_threshold = 22)  # e.g. fraction of steps equatorward
plot_probgls_track(batch, colony = cfg$tagging.location)
```

Each deployment is processed independently with robust error handling, and a
per-deployment track CSV plus a `pipeline_summary.csv` are written to
`output_dir`.

## Reproducibility

* The `config.yml` fully specifies the movement model (29 `prob_algorithm`
  parameters) — archive it with your results.
* `run_probgls()` passes only the arguments the installed `probGLS` declares, so
  analyses remain reproducible across probGLS versions.
